---
title: 'DS612Group Project Fall 2020: Phase 1'
author: "Jimmy Wen, Joel Gumm, Seth Hill, Tatiana Macha, Vivienne Zhang"
date: "10/6/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Cleaning
We began the data cleaning process by dropping NA values from the data set. From there, we dropped uneccesary columns (id, lat, long) and began searching for outliers by calculating some basic statistics. 
```{r}

#Load data
data = read.csv("DS 612 Group Project data set.csv")

#omit na values 
data = na.omit(data)

#remove unnecessary columns
data <- data[-c(1, 18, 19)]

#Define quant and qual variables
quants <- subset(data, select = -date)
quals <- data$date

#Outlier detection
multi_func <- function(x){
  c(min = min(x), mean = mean(x), max = max(x), std_dev = sd(x))
}

sapply(quants, multi_func)
#Investigate further: price, bedrooms, yr_renovated
```
From this, we found that the variables price and bedrooms required further investigation. 
 
``` {r}
#Price
hist(data$price, main = "Histogram of Price", xlab = "Price")
```
We plan on using price as our dependent variable in our linear regression problem. After looking at its distribution, we can see that some transformation is required in order to normialze its distribution and make it a worth dependent variable. 

``` {r}
#logarithmic transformation to fix right skew in outcome variable
data['log_price'] <- log(data['price'])
```

After this transformation, the variable is now normally distributed. 

``` {r}
hist(data$log_price, main = "Histogram of log_price", xlab = "log_price") #normal
#Redefine quants to include LogPrice
quants <- subset(data, select =-date)

```

Moving on to bedrooms, when we ran the summary statistics, we could see that there were observations that included houses with 33 bedrooms. Even if houses like these do exist, they are an extreme outlier and should be dropped from our dataset.

``` {r}
which(data$bedroom == 33)
data <- data[-c(15871), ]
```


After we handled our outliers, we created dummy variables on the columns we believed would make for satisfactory categorical variables and converted the data variable from char to date type. 

``` {r}
data["popular"] <- ifelse(data$view >2, 1, 0)
data["good_condition"] <- ifelse(data$condition >= 3, 1, 0)
data["high_grade"] <- ifelse(data$grade >= 7, 1, 0)

#Convert dates 
data$date <- sapply(data$date, substring, 1, 8)
data$date <- as.Date(data$date, "%Y%m%d", tz = tz)
```

Finally the cleaned data was written into a .csv file for future use. 
``` {r}
write.csv(data, file = "clean_data.csv", col.names = TRUE)
```

