---
title: 'DS612Group Project Fall 2020: Phase 1'
author: "Jimmy Wen, Joel Gumm, Seth Hill, Tatiana Macha, Vivienne Zhang"
date: "10/6/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Cleaning
We began the data cleaning process by dropping NA values from the data set. From there, we dropped uneccesary columns (id, lat, long) and began searching for outliers by calculating some basic statistics. 
```{r}

#Load data
data = read.csv("DS 612 Group Project data set.csv")

#omit na values 
data = na.omit(data)

#remove unnecessary columns
data <- data[-c(1, 18, 19)]

#Define quant and qual variables
quants <- subset(data, select = -date)
quals <- data$date

#Outlier detection
multi_func <- function(x){
  c(min = min(x), mean = mean(x), max = max(x), std_dev = sd(x))
}

sapply(quants, multi_func)
#Investigate further: price, bedrooms, yr_renovated
```
From this, we found that the variables price and bedrooms required further investigation. 
 
``` {r}
#Price
hist(data$price, main = "Histogram of Price", xlab = "Price")
```
We plan on using price as our dependent variable in our linear regression problem. After looking at its distribution, we can see that some transformation is required in order to normialze its distribution and make it a worth dependent variable. 

``` {r}
#logarithmic transformation to fix right skew in outcome variable
data['log_price'] <- log(data['price'])
```

After this transformation, the variable is now normally distributed. 

``` {r}
hist(data$log_price, main = "Histogram of log_price", xlab = "log_price") #normal
#Redefine quants to include LogPrice
quants <- subset(data, select =-date)

```

Moving on to bedrooms, when we ran the summary statistics, we could see that there were observations that included houses with 33 bedrooms. Even if houses like these do exist, they are an extreme outlier and should be dropped from our dataset.

``` {r}
which(data$bedroom == 33)
data <- data[-c(15871), ]
```


After we handled our outliers, we created dummy variables on the columns we believed would make for satisfactory categorical variables and converted the data variable from char to date type. 

``` {r}
data["popular"] <- ifelse(data$view >2, 1, 0)
data["good_condition"] <- ifelse(data$condition >= 3, 1, 0)
data["high_grade"] <- ifelse(data$grade >= 7, 1, 0)

#Convert dates 
data$date <- sapply(data$date, substring, 1, 8)
data$date <- as.Date(data$date, "%Y%m%d", tz = tz)
```

Finally the cleaned data was written into a .csv file for future use. 
``` {r}
write.csv(data, file = "clean_data.csv", col.names = TRUE)
```

## Descriptive Statistics
With the dataset cleaned, we proceeded to describe the data for gathering insights and model building.

Starting with summarizing the cleaned dataset:

``` {r}
#summary
summary(data)
```

Categorical variables like properties presiding on waterfronts, being regarded as a high grade property, or the zipcode in which the property is located can largely be disregarded when evaluating the above summary table.  The importance of this table can be found in the high level overview it provides on variables such as year built, condition, different iterations of square feet, and the presence of properties regarded as high grade.  At a glance we can see a snapshot of our dataset that shows the mean construction year of houses across all markets is 1971, and the mean of properties considered to be high grade (>7) is .89, or 89% of observed properties are considered high grade properties.  Being that our primary function is to show how different factors affect housing prices, we will want to dive further into descriptive methods for directing our models.

Starting next by evaluating a Correlation Matrix of our variables to identify relationships:
``` {r}
#correlation matrix of cleaned data

corr <- round(cor(data()),2)
corrplot(corr,method ="circle")
```

The high correlation between log price and price can be disregarded due to its function being for normality.*

From the start we can see that price has the highest positive correlation with the square feet of the living area for a property. For every one digit increase in square feet of the living area, we could expect a 0.7 increase in the price of the property. The grade of the property also has a relatively high positive correlation, with a value of 0.67. By no means conclusive on what to use as predictors, these moderately high correlations give us hints of where we should direct our initial investigations.

Since Grade is a discrete variable, it lends itself more easily to gathering insights from reviewing the frequency of different property grades.
``` {r}
#Frequency Table for Grade

t <- as.data.frame(table(data$grade))
names(t)[1]='grade'
t

```
This is information which simply using a correlation matrix could not provide. The majority of observations lie between bins 7 and 8 for grade; making the vast majority of properties high grade properties, but still very close to the cutoff. It should come as no surprise that as Grade increases, the amount of observations sharply decline.

Now we will look at the average prices for the different Grade groups.
``` {r}
#Average price per grade class 

a <- aggregate(data$price ~ data$grade, FUN = mean)
names(a)[1] = "Grade"
names(a)[2] = "Avg Price"
a

```

As could be expected, price increases as Grade increases, which was observed in the correlation matrix. The differences for every grade increase sub 7 is quite marginal. After arriving at high grade properties, which are 7 and above, an exponential price increase can be observed.

For continuous variables, the relationship to price can be better understood with the scatter plots below.
``` {r}
#Sqft_living vs Avg price
plot(data$sqft_living, data$price)
#sqft_basement class
plot(data$sqft_basemen, data$price)
```
Living squarefeet increases with price, but the spread also increases. 

While minor changes in price as basement square feet increase can be observed, it would at most be a weak positive correlation. 


``` {r}
#Histogram of grade 

#histogram of sqft_living
hist(data$sqft_living, 
     main = "Histogram of Sqft_Living",
     xlab = "Housing Sqft_Living",
     ylab = "Frequency")

```
This histogram shows more succinctly the distribution of living space sizes. The largest group falling in the 1500-3000 sqft range. 


Viewing this relationship from a distributions view, we see the spread of the data increases dramatically once you reach the high grade level. 
``` {r}
plot(factor(data$grade),data$price,
     main = "Box Plot of Price vs. Grade",
     xlab = "Grade",
     ylab = "Price")
```